\textbf{ Matching the forecast horizon with the relevant ecological processes }

### 2019-07-22 doc/manuscript.tex
Editorial work on Eco-Evo example

We compare forecasts from both models to the true, simulated dynamics to show that responses to environmental change reflect both kinds of processes, but at different time scales. To quantify this pattern, we combine forecasts from the spatial and temporal approaches with a weighted average and then show how the optimal model weights change over time: we expect the temporal model to best predict short-term dynamics, and the spatial model to best predict long-term dynamics. How rapidly that transition occurs depends on the rate of  ``slow" processes such as species turnover and evolution. 
Code for simulating dynamics and reproducing the analyses in this paper is available on Github \texttt{https://github.com/pbadler/space-time-forecast}) and
archived on Zenodo \texttt{https://zenodo.org}).

@@

A common need for conservation biologists and natural resource managers is anticipating the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just the focal species, 
skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do
altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018}. 
The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient.
Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: 
cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by 
fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. 
Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability.

@@

During the baseline period there were strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our ``spatial model" of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature (Fig. \ref{fig:species-patterns-models}B). Our ``temporal model" estimates this weak, linear temperature effect, along with strong lag effects of biomass in the previous year. The spatial and temporal statistical
models are described in Appendix \ref{models}.

@@

We forecast the effect of a temperature increase on the focal species' biomass at one location in the center of the temperature gradient 
using both spatial models and the temporal models. 
The predictions from the spatial and temporal models contrasted markedly, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease.Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the focal site, the biomass of the focal species decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}).

To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix \ref{models} for full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}), but 
it works so effectively because the weights were determined by fitting directly to those data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from being dominated by interannual 
variation captured in the temporal model to being dominated by the steady-state equilibrium captured by the spatial model.
A true, combined forecast would require a method to determine the model weights \emph{a priori}.
The most rapid shifts in the model weights occurred during the period when warm-adapted, faster growing species were increasing most rapidly
in abundance (Fig. \ref{fig:community-weights-spp}).

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location (refs). Like the shifts
in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales.
We demonstrate this process with a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). The full model description is provided in Appendix \ref{eco-evo}, and all computer code is available on Github  (\texttt{https://github.com/pbadler/space-time-forecast}). Genotype frequencies will shift across a gradient of mean annual temperature: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B).

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We used the simulated data from this baseline period to fit an empirical model that assumes no knowledge of the underlying eco-evolutionary process. This empirical ``temporal model" (Appendix \ref{models}) predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:evo_evo_spatial_temporal_models}C).

We then imposed a period of warming, followed by a final period of higher, but stationary, temperature (Fig. \ref{fig:forecast} top). With the onset of warming, the population crashed as the cold-adapted genotype decreased in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype began to increase and the population recovered (Fig. \ref{fig:forecast} bottom). The implications for forecasting are clear: In the initial stages of the warming trend, the temporal model fit to the original, baseline conditions should make better the best predictions about resulting population trends, but in the longer term, as evolutionary change occurs, the spatial model ultimately makes better predictions.

As in the eco-evolutionary example, we also fit a combined model as a weighted average of the spatial and temporal model, with the weights changing over time. 
The temporal model accurately predicted the impact of the initial warming trend, but eventually became far too pessimistic, while the spatial model did not handle the initial trend but accurately predicted the eventual, new steady state (Fig. \ref{fig:forecast} bottom). The combined model initially reflected the temporal model, but then rapidly transitioned to reflect the spatial model. The rapid transition in the weighting term, $\omega$, occured during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}). The combined model's predictions look impressively accurate, but, as in the community turnover example, that is because we used the full, simulated time series to fit the weighting term. 
To generate a true forecast from the combined model, we would need an independent method to predict how the model weights shift over time.

>>>>

We compare forecasts from both models to the true, simulated dynamics to show that responses to environmental change reflect both kinds of processes, but at different time scales. To quantify this pattern, we combine forecasts from the spatial and temporal approaches with a weighted average and then show how the optimal model weights change over time: we expect the temporal model to best predict short-term dynamics, and the spatial model to best predict long-term dynamics. How rapidly that transition occurs depends on the rate of  ``slow" processes such as species turnover and evolution.

@@

A common problem for conservation biologists and natural resource managers is anticipating the impact of environmental change on the abundance of 
endangered species, biological invaders, and harvested species. Although the managers may be primarily interested in just the focal species, 
skillful prediction might require considering interactions with many other species, greatly complicating the problem. But at what forecast horizon do
altered species interactions become impossible to ignore? We explored this question using a metacommunity model developed by \cite{alexander_lags_2018}. 
The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient.
Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: 
cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by 
fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. 
Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability. All computer code is available on Github: \texttt{https://github.com/pbadler/space-time-forecast}).

@@

Dynamics during the baseline period revealed very strong spatial patterns across the mean temperature gradient. 
Individual species, including our focal species, showed classic, unimodal ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:species-patterns-models}A).
These spatial patterns are the basis for our ``spatial model" of the temperature-biomass relationship for our focal species (Fig. \ref{fig:species-patterns-models}A). 
In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, the biomass of the focal species was quite insensitive to interannnual variation in temperature (Fig. \ref{fig:species-patterns-models}B). Our ``temporal model" estimates this weak, linear temperature effect, along with strong lag effects of biomass in the previous year. The spatial and temporal statistical
models are described in Appendix \ref{models}.

@@

We forecasted the effect of temperature increase on the focal species' biomass at one location in the center of the temperature gradient 
using the spatial models and the temporal models. 
The predictions from the spatial and temporal model contrasted markedly, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease. Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species colonized and increased in abundance at the study site, the biomass of the focal species then decreased, eventually falling below its baseline level (Fig. \ref{fig:community-forecast-species}).

To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix \ref{models} for full description of the approach).
This combined model accurately predicts the simulated dynamics across the full forecast horizon (Fig. \ref{fig:community-forecast-species}) 
because the weights were determined by fitting directly to those data. Unlike our spatial and temporal model forecasts, 
we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from from being dominated by the interannual 
variations captured in the temporal model to being dominated by the steady-state equilibrium captured by the spatial model.
The most rapid shifts in model weights occur during the period when warm-adapted, faster growing species are increasing most rapidly
in abundance (Fig. \ref{fig:community-weights-spp}). A true, combined forecast would require a method to determine the model weights \emph{a priori}.

@@

Evolutionary adaptation is a key uncertainty in predicting how environmental change will impact a focal population at a given location (refs). Like the shifts
in species composition illustrated in the previous example, shifts in genotype frequencies can also influence dynamics and forecasts at different time scales.
We demonstrate this process with a hypothetical annual plant population in which fecundity is temperature dependent, and different genotypes have different temperature optima (Fig. \ref{fig:evo_evo_spatial_temporal_models}A). The full model description is provided in Appendix \ref{eco-evo}. Genotype frequencies will shift across a gradient of mean annual temperature: cold sites will be dominated by the cold-adapted homozygous genotype, warm sites will be dominated by the heat-adapted homozygous genotype, and intermediate sites will be dominated by the heterozygous genotype (Fig. \ref{fig:evo_evo_spatial_temporal_models}B).

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We used the simulated data from this baseline period to fit an empirical model that assumes no knowledge of the underlying eco-evolutionary process. This empirical ``temporal model" (Appendix \ref{models}) predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:evo_evo_spatial_temporal_models}C). We then imposed a period of warming, followed by a final period of higher stationary temperature (Fig. \ref{fig:forecast} top).

With the onset of warming, the population crashed as the cold-adapted genotype decreased in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype began to increase and the population recovered (Fig. \ref{fig:forecast} bottom). The temporal model accurately predicted the impact of the initial warming trend, but eventually became too pessimistic, while the spatial model did not handle the initial trend but accurately predicted the eventual, new steady state (Fig. \ref{fig:forecast} bottom).

As in the community turnover example, we also fit a combined model as a weighted average of the spatial and temporal model, with the weights changing over time. The combined model initially reflected the temporal model, but then rapidly transitioned to reflect the spatial model. The rapid transition in the weighting term, $\omega$, occurred during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}). The combined model's predictions look impressively accurate, but, as in the community turnover example, that is because we used the full, simulated time series to fit the weighting term. To generate a true forecast from the combined model, we would need an independent method to predict how the model weights shift over time.
